Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-07-18 Manuscript/manuscript.tex
saving work

The above definition ensures that all types of determinants with a given scaling enter at the same hierarchy $h$.
% computational scaling
We show in Fig.~\ref{fig:hCI} how the seniority-excitation map of determinants is partitioned for a closed-shell reference, and for open-shell references with one or two unpaired electrons.
For these three types of reference, Fig.~\ref{fig:determinants} shows the classes of determinants included up to hCI1.
hCI can also be interpreted as the CI model that provides increasingly more dissimilar determinants from the given reference determinant.
This dissimilarity is represented by the hierarchy $h$, which accounts for differences in orbital occupation (through the excitation degree $e$)
and differences in the number of unpaired electrons (through the term $(s-s_0)/2$).

@@

The first order for a closed-shell reference determinant is hCI1 (see top left of Fig.~\ref{fig:determinants}), which accounts for all single excitations (as CIS) 
plus all paired doubled excitations (two electrons promoted from the same occupied orbital into the same virtual orbital).
For both types of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, thus the empty blocks at $s=0$ and $e$ odd.
hCI1.5 includes to the hCI1 determinants the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 accounts for all types of determinants where $\Ndet$ scale as $N^4$ or less.
That means all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons, 
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Similarly, for excitation-based CI, $\Ndet$ scales as $N^{2e}$, where $e$ is the excitation degree.

@@

Finally, one can employ a high-spin reference determinant to define the hCI models.
The case of two unpaired electrons is shown in the right pannel of Fig.~\ref{fig:hCI}.
The key difference with respect to the closed-shell case is that the hCI sequence is displaced one block to the right in the seniority-excitation map,
reflecting the shift from the $s_0=0$ to the $s_0=2$ reference.
In the two previous cases, the hCI0 level only includes the reference determinant.
For a $s=2$ reference, however, hCI0 not only accounts for the reference determinant,
but also the two closed-shell determinants produced by the single excitation that pairs the two unpaired electrons.
The lower order classes of determinants can be seen in the bottom of Fig.~\ref{fig:determinants},
which clearly outnumber the fewer classes associated with a closed-shell reference.
One could adopt even higher-spin references, further displacing the hCI sequence to the right.
In this case, hCI0 would include all the lower spin determinants obtained by partially or totally pairing the unpaired electrons.
Notice that the number of such determinants do not depend on the size of the system (scaling as $N^0$), and are thus included at the $h=0$ level, in line with the spirit of hCI.
hCI can be built in a similar way on top of a high-spin reference for odd number of electrons.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space defined by the reference determinant(s) and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the CI calculations presented here.
In a given calculation, the energies are considered to be converged when the (largest) renormalized EN2 correction computed in the truncated Hilbert space 
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
%by spanning the most important regions of the truncated Hilbert space
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delievering fairly converged absolute and excitation energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we futher evaluated the bare and the renormalized EN2 perturbative correction. \cite{Garniron_2019} 
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for excitation-based CI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.

@@

The results were analyzed along the same lines of our report on hCI for closed-shell systems. \cite{Kossoski_2022}
Namely, for the different CI models considered here, 
we evaluated the convergence of the non-parallelity error (NPE), the distance error, the harmonic vibrational frequencies, and the equilibrium bond lengths, as functions of $\Ndet$.
The NPE of a given level of theory is defined as the maximum minus the minimum differences between its corresponding PEC and the FCI one, for a given range of coordinates.
Here we redefine the previous definition of the distance error \cite{Kossoski_2022} to accommodate for the fact the approximate PEC might appear below the FCI when perturbative corrections are employed.
The previous definition could provide negative values in this case, whereas this metric should provide the overall extent of proximity between two PECs.
Here, we define the distance error based on the signed differences between two PECs, as the absolute value of their maximum difference plus the absolute value of their minimum difference, 
evaluated at given coordinate interval.
Details about how the harmonic vibrational frequencies and equilibrium geometries were obtained from the calculated PECs,
along with the ranges defining the NPE and distance errors can be found in the \SupInf.

@@

% vinyl missing point at 5.2 for s3i+pt2 and at 2.0 for s5+pt2
% CN correct dissociating part for h4 and h3.5
% H8 missing points for Q+pt2
% H8 missing 9.0 for T+pt2

The full set of PECs for the open-shell systems (\ce{OH}, \ce{CN}, \ce{H7}, and vinyl) are presented in the {\SupInf},
whereas Fig.~\ref{fig:plot_all} contains the NPEs, distance errors, equilibrium bond lengths and harmonic vibrational frequencies plotted as a function of $\Ndet$ for the different CI models.
The present results for open-shell systems display similar trends to the ones previously reported for closed-shell systems. \cite{Kossoski_2022}
The main finding is that hCI typically improves the convergence of the different properties when compared to excitation-based CI.
Concerning the NPEs, such improvement is clear for \ce{OH} and \ce{CN} (involving a single bond breaking), becoming less so for vinyl (double bond breaking),
whereas for \ce{H7} (multiple bond breaking), hCI and excitation-based CI perform similarly.
This dependence on the order of bond breaking had also been observed for closed-shell systems. \cite{Kossoski_2022}
The same trend is observed for the distance error, although the improvement from excitation-based CI to hCI is less marked than for the NPEs.
hCI further leads to a faster convergence of the equilibrium geometry of \ce{CN}, but does not improve upon excitation-based CI for the geometries of the other systems.
Finally, hCI systematically outperforms excitation-based CI in the calculation of vibrational frequencies, except for \ce{H7}, where no big difference is found.

%%% FIG 3 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all}
\caption{
Non-parallelity errors, distance errors, equilibrium distances, and vibrational frequencies (or force constants), for \ce{OH}, \ce{CN}, \ce{H7}, and vinyl,
as a function of the number of determinants, according to hCI (green), excitation-based CI (red) and sCI (blue) models.}
\label{fig:plot_all}
\end{figure*}
%%% %%% %%%

sCI models deliver a poor convergence of all observables.
The single exception is the NPE for \ce{OH} obtained from sCI0, smaller than the NPEs obtained from CISD and hCI1.5 or hCI2, methods having a comparable computational cost in this case.
Going to larger basis sets or larger systems, the computational burden of sCI models would increase considerably more than hCI or excitation-based CI models, 
due to its exponential scaling, rendering this route even less atractive.
Analogous conclusions had been drawn for sCI applied to closed-shell systems, \cite{Kossoski_2022}
and here we show that the same holds for radicals.

@@

We further assesssed the effect of the EN2 perturbative correction for both the open-shell systems studied here 
and the closed-shell systems (\ce{HF}, \ce{F2}, \ce{N2}, ethylene, \ce{H4}, and \ce{H8}) addressed in our first report on hCI. \cite{Kossoski_2022}
The corresponding PECSs are shown in the {\SupInf}.
At stretched geometries, the perturbative correction can lead to kinks and large jumps in the PECs.
Such problems were encountered at some levels of CI for \ce{CN}, vinyl, \ce{H8}, and \ce{H7}, and to a less extent, for \ce{N2} (only for hCI1).
For the remaining five systems (\ce{HF}, \ce{F2}, ethylene, \ce{H4}, and \ce{OH}),
the EN2 correction produced smooth PECs at dissociation.
Around the equilibrium geometry, the perturbative correction produced well-behaved PECs at all levels of CI and for all systems, 
except for \ce{CN}.
In this case, there are several close-lying excited states around the equilibrium geometry. \cite{}
Since the unperturbed PEC corresponds to the lowest-lying solution, it often presents kinks where the ground state changes its character.
Because of that, the perturbative correction produces discontinuous PECs.

%%% FIG 4 %%%
\begin{figure}%[h!]
\includegraphics[width=1.0\linewidth]{plot_oh_rpt2}
\includegraphics[width=1.0\linewidth]{plot_f2_rpt2}
\caption{
Non-parallelity errors, distance errors, equilibrium distances, and vibrational frequencies, for \ce{OH} (top four pannels) and \ce{F2} (bottom four pannels),
as a function of the number of determinants, according to hCI (green), excitation-based CI (red) and sCI (blue) models,
with (light tones) and without (dark tones) the EN2 perturbative correction.
}
\label{fig:plot_rpt2}
\end{figure}
%%% %%% %%%

To illustrate the effect of the EN2 perturbative correction, we show the results for \ce{OH} and \ce{F2} in Fig.\ref{fig:plot_rpt2}.
The full set of results can be found in the \SupInf.
We found that the EN2 correction generally reduces the errors of all observables considered here, for a given level of CI.
As such, it accelerates the rate of convergence as a function of the number of determinants in the unperturbed calculation.
Importantly, the perturbative corrected hCI route still outperforms the excitation-based counterpart.
Furthermore, the reduction of the errors is overall more monotonic than without the perturbative correction,
clearly seen in the case of \ce{F2}, shown in Fig.\ref{fig:plot_rpt2}.
In some cases, the EN2 correction does not improve the convergence of the sCI route.
This can be seen for the equilibrium geometry and vibrational frequency of \ce{F2}.

%sCI route, however, the EN2 correction does not bring a clear improvement to render it competitive in relation to the excitation-based and hCI alternatives.

@@

@@

For each CI model considered here, we evaluated the mean signed error (MSE), mean absolute error (MAE), and root-mean square error (RMSE) 
with respect to reference theoretical values for the excitation energies.
The statistical errors are presented in Tab.~\ref{tab:1} and Tab.~\ref{tab:2},
for singly excited states of closed-shell and open-shell systems, respectively.
For the lower order models, the calculations were performed for 46 (closed-shell) and 20 (open-shell) excited states.
In turn, a subset of 14 (closed-shell) and 5 (open-shell) excited states was considered for the higher order models, given their more intensive computational cost.
Even though these subsets are too small for meaningful absolute statistics, they can still show the main trends on the higher order CI models.
%A detailed comparison of excitation-based CI along the ground-state and the state-specific routes can be found elsewhere. \cite{Kossoski_2023}
%The focus of the present discussion lies on the comparison between hCI, sCI, and excitation-based CI, for both routes,

@@

We start the discussion with the excitations from closed-shell systems, with corresponding statistical errors shown in Tab.~\ref{tab:1}.
%
The first level of hCI, hCI1, produces poor excitation energies, with a MAE of \SI{1.07}{\eV}.
This model shares the same computational scaling with CIS, but includes the paired double excitations, 
which clearly do not improve the descent CIS results (MAE of \SI{0.61}{\eV}).
Moving one rank up, to hCI1.5, the MAE increases to \SI{1.91}{\eV}, and reaches a maxium of \SI{3.53}{\eV} at the hCI2 level.
Here we notice that hCI2 delivers somewhat smaller errors than CISD (MAE of \SI{4.09}{\eV}), even though both are way too large.
From that point on, the next hCI models generate progressively better results.
Despite the improvement with respect to hCI2, hCI2.5 is stil as inaccurate (MAE of \SI{1.95}{\eV}) as hCI1.5.
At the hCI3 level, significantly smaller errors are finally achieved (MAE of \SI{0.19}{\eV}).
This is quite close to that obtained with the excitation-based model of same order, CISDT (\SI{0.17}{\eV}).
Even smaller errors are produced at the hCI3.5 level (MAE of \SI{0.11}{\eV}), but at a considerable computational cost.

@@

%we further investigated whether higher-order versions could be competitive against the traditional excitation-based counterparts.
%For that, we performed calculations for a subset of 14 singly-excited states of small closed-shell molecular systems (indicated in the {\SupInf}).
%(\ce{H2O}, \ce{BF}, \ce{HCF}, \ce{HCl}, and \ce{N2}),
%with corresponding statistical errors shown in Tab.~\ref{tab:1}.
%Even though this is a small set, it should be enough to obtain the main trends on the accuracy of higher-order models.

@@

This residual bias disappears at the next order, $\Delta$hCI2, whereas for the closed-shell systems even $\Delta$hCI3.5 still understimated excitation energies. 
The same holds for excitation-based CI.
For open-shells, $\Delta$hCI2 is arguably slightly more accurate (MAE of \SI{0.11}{\eV}) than $\Delta$CISD (MAE of \SI{0.12}{\eV}), also based on the larger differences on the RMSEs.
In further contrast to the closed-shell case,
$\Delta$hCI2.5 is more accurate than $\Delta$hCI2, producing a MAE of \SI{0.06}{\eV}.
(Accounting only for the 5 states considered for $\Delta$hCI2.5, $\Delta$hCI2 would have a MAE of \SI{0.09}{\eV}.)
The statistics at the next orders ($h=3$ and $h=3.5$) is limited, 
but suggests a possible advantage of the hCI models against the excitation-based ones,
and that the ground-state routes perform slightly better than the state-specific ones.

>>>>

The above definition ensures that all types of determinants with a given scaling enter at the same hierarchy $h$.
% computational scaling
We show in Fig.~\ref{fig:hCI} how the seniority-excitation map of determinants is partitioned for a closed-shell reference, and for open-shell references with one or two unpaired electrons.
For these three types of reference, Fig.~\ref{fig:determinants} shows the classes of determinants included up to hCI1.
hCI can also be interpreted as the CI model that provides increasingly more dissimilar determinants from the given reference determinant.
The level of dissimilarity is represented by the hierarchy $h$, which accounts for differences in orbital occupation (through the excitation degree $e$)
and differences in the number of unpaired electrons (through the term $(s-s_0)/2$).

@@

The first order for a closed-shell reference determinant is hCI1 (see top left of Fig.~\ref{fig:determinants}), which accounts for all single excitations (as CIS) 
plus all paired doubled excitations (two electrons promoted from the same occupied orbital into the same virtual orbital).
For both types of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, thus the empty blocks at $s=0$ and $e$ odd.
hCI1.5 includes to the hCI1 determinants the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 accounts for all types of determinants where $\Ndet$ scale as $N^4$ or less.
That means all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons, 
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Similarly, for excitation-based CI, $\Ndet$ scales as $N^{2e}$, where $e$ is the excitation degree.
The computational scaling along the two routes is shown in Tab.~\ref{tab:scaling}.

\begin{table}[ht!]
\caption{Computational scaling along the hierarchy-based and excitation-base CI routes, with and without the EN2 perturbative correction.}
\label{tab:scaling}
\begin{ruledtabular}
\begin{tabular}{ll|ll|l}
 \mc{2}{c|}{hierarchy-based}   &  \mc{2}{c|}{excitation-based} &  scaling \\
\hline
hCI1   &    -       & CIS    &   -       & $\order*{N^4}$ \\
hCI1.5 &    -       &  -     &   -       & $\order*{N^5}$ \\
\hline
hCI2   & hCI1+EN2   & CISD   & CIS+EN2   & $\order*{N^6}$ \\
hCI2.5 & hCI1.5+EN2 &  -     &   -       & $\order*{N^7}$ \\
\hline
hCI3   & hCI2+EN2   & CISDT  & CISD+EN2  & $\order*{N^8}$ \\
hCI3.5 & hCI2.5+EN2 &   -    &   -       & $\order*{N^9}$ \\
\hline
hCI4   & hCI3+EN2   & CISDTQ & CISDT+EN2 & $\order*{N^{10}}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

@@

Finally, one can employ a high-spin reference determinant to define the hCI models.
The case of two unpaired electrons is shown in the right pannel of Fig.~\ref{fig:hCI}.
The key difference with respect to the closed-shell case is that the hCI sequence is displaced one block to the right in the seniority-excitation map,
reflecting the shift from the $s_0=0$ to the $s_0=2$ reference.
In the two previous cases, the hCI0 level only includes the reference determinant.
For a $s=2$ reference, however, hCI0 not only accounts for the reference determinant,
but also the two closed-shell determinants produced by the single excitation that pairs the two unpaired electrons.
The lower order classes of determinants can be seen in the bottom of Fig.~\ref{fig:determinants},
which clearly outnumber the fewer classes associated with a closed-shell reference.
One could adopt even higher-spin references, further displacing the hCI sequence to the right in the seniority-excitation map.
In this case, hCI0 would include all the lower spin determinants obtained by partially or totally pairing the unpaired electrons.
Notice that the number of such determinants do not depend on the size of the system (scaling as $N^0$), and are thus included at the $h=0$ level, in line with the spirit of hCI.
Similarly, hCI can be built on top of a high-spin reference for odd number of electrons.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space defined by the reference determinant(s) and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the CI calculations presented here.
In a given calculation, the energies are considered to be converged when the (largest) renormalized EN2 correction computed in the truncated Hilbert space 
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
%by spanning the most important regions of the truncated Hilbert space
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delievering fairly converged absolute and excitation energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we futher evaluated the bare and the renormalized EN2 perturbative correction. \cite{Garniron_2019} 
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for excitation-based CI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.
The computational scaling associated with the CI+EN2 calculations is also presented in Tab.~\ref{tab:scaling}.

@@

The results were analyzed along the same lines of our report on hCI for closed-shell systems. \cite{Kossoski_2022}
Namely, for the different CI models considered here, 
we evaluated the convergence of the non-parallelity error (NPE), the distance error, the harmonic vibrational frequencies, and the equilibrium bond lengths, as functions of $\Ndet$.
The NPE of a given level of theory is defined as the maximum minus the minimum differences between its corresponding PEC and the FCI one, for a given range of coordinates.
Here we redefine the previous definition of the distance error \cite{Kossoski_2022} to accommodate for the fact the approximate PEC might appear below the FCI when perturbative corrections are employed.
The previous definition could provide negative values in this case, whereas this metric should provide the overall extent of proximity between two PECs.
Here, we define the distance error based on the signed differences between two PECs, as the absolute value of their maximum difference plus the absolute value of their minimum difference, 
evaluated at given coordinate interval.
In the discussion, we say equilibrium properties when referring to both the equilibrium geometry and the vibrational frequency.
Details about how the harmonic vibrational frequencies and equilibrium geometries were obtained from the calculated PECs,
along with the ranges defining the NPE and distance errors can be found in the \SupInf.

@@

% vinyl eq.geom e4 s5 h3.5 rpt2
% vinyl 5.0 h3+rpt2
% update pecs of ethylene

The full set of PECs for the open-shell systems (\ce{OH}, \ce{CN}, \ce{H7}, and vinyl) are presented in the {\SupInf}.
From these PECs, we obtain the NPEs, distance errors, equilibrium bond lengths and harmonic vibrational frequencies, which are plotted as a function of $\Ndet$ in Fig.~\ref{fig:plot_all}.
The corresponding results for the closed-shell systems (\ce{HF}, \ce{F2}, \ce{N2}, ethylene, \ce{H4}, and \ce{H8}) are shown in Fig.~\ref{fig:plot_all_closed}.
When we first introduced hCI, \cite{Kossoski_2022} we surveyed the same molecules, though only for the bare CI models (without a perturbative correction).
These previous results are reproduced in Fig.~\ref{fig:plot_all_closed} together with the present results, which include such correction.
First we will focus on the results for the bare CI models (hCI, excitation-based CI and sCI) for the open-shell systems, represented by the dark tones,
leaving the discussion about the perturbative correction (light tones) for later.

%%% FIG 3 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{OH}, \ce{CN}, \ce{H7}, and vinyl,
as a function of the number of determinants ($\Ndet$), according to hCI (green), excitation-based CI (red) and sCI (blue) models,
with (light tones) and without (dark tones) the EN2 perturbative correction.
The dashed lines represent the FCI results.}
\label{fig:plot_all}
\end{figure*}
%%% %%% %%%

We found that hCI typically improves the convergence with respect to $\Ndet$ when compared to excitation-based CI, for the four properties surveyed here.
The present finding for open-shell systems therefore parallels the previously reported superiority of hCI for closed-shell systems. \cite{Kossoski_2022}
Concerning the NPEs, the advantage of hCI is pronounced for \ce{OH} (involving a single bond breaking), 
becoming less so, though still clear, for vinyl (double bond breaking) and \ce{CN} (triple bond breaking),
whereas for \ce{H7} (multiple bond breaking), it still shows a small advantage in comparison to excitation-based CI.
This dependence on the order of bond breaking had also been observed for closed-shell systems \cite{Kossoski_2022} (as can be seen in Fig.~\ref{fig:plot_all_closed}).
We also found overall smaller distance errors with hCI than with excitation-based CI, although the improvement is less noticeable in this case.
hCI further leads to a somewhat faster convergence of the equilibrium geometries for the open-shell systems, most noticeably for \ce{CN}.
Finally, hCI systematically outperforms excitation-based CI for the calculation of vibrational frequencies, except for \ce{H7}, where no big difference is found.

In great contrast to hCI (and excitation-based CI),
sCI models deliver a poor convergence of all observables.
The single exception is the NPE for \ce{OH} obtained from sCI0, smaller than the NPEs obtained from CISD and hCI1.5 or hCI2, methods having a comparable computational cost in this particular case.
Going to larger basis sets or bigger systems, the computational burden of sCI models would increase considerably more than hCI or excitation-based CI models, 
due to its formal exponential scaling.
In their own, we conclude that the sCI route is unatractive to open-shell systems,
in line with similar findings for closed-shell systems \cite{Kossoski_2022} (also clear from Fig.~\ref{fig:plot_all_closed}).

@@

The results for the CI models corrected by the EN2 energy are represented by the light tones in Figs.~\ref{fig:plot_all} and ~\ref{fig:plot_all_closed}.
%The corresponding PECSs are shown in the {\SupInf}.
We excluded cases where there are important discontinuities in the PECs.
These discontinuities reflect crossings between states of different symmetries in the unperturbed calculation,
appearing as kinks in the computed ground-state PEC.
Due to the abrupt change of character, the EN2 correction is not uniform and gives rise to the observed discontinuities.
This happens for the open-shell systems at dissociation (except for \ce{OH}), for which we do not present NPE and distance errors.
For \ce{OH} and the six closed-shell systems, the EN2 correction produced smooth PECs at dissociation.
Around the equilibrium geometry, the EN2 correction also leads to well-behaved PECs, for all CI models and for all systems, except for \ce{CN}
(thus no equilibrium properties are presented for this radical).
This is simply due to its several close-lying excited states. \cite{}
If we were to follow the PEC of a given symmetry (and not the lowest-lying state as we did here), there would be no discontinuity stemming from the EN2 correction.

%%% FIG 4 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{plot_all_closed}
\caption{
Non-parallelity error (NPE), distance error, equilibrium distance, and vibrational frequency (or force constant), for \ce{HF}, \ce{F2}, ethylene, \ce{N2}, \ce{H4}, and \ce{H8}, 
as a function of the number of determinants ($\Ndet$), according to hCI (green), excitation-based CI (red) and sCI (blue) models,
with (light tones) and without (dark tones) the EN2 perturbative correction.
The dashed lines represent the FCI results.}
\label{fig:plot_all_closed}
\end{figure*}
%%% %%% %%%

We found that the EN2 correction significantly reduces the errors with respect to the FCI results.
This is observed for all the CI models, systems, and observables considered here.
In the particular case of \ce{OH}, the improvement is massive, as even the lower levels of CI+EN2 deliver very close results to FCI.
The effect of the EN2 correction is most noticeable for the distance error and equilibrium properties, and less so for the NPE.
In few cases, however, it does not improve the bare CI results, suggesting substantial cancelation of errors in the latter.
This is seen at the hCI1 and hCI1.5 models and their EN2 corrected couterparts, for the NPE of \ce{HF}, \ce{F2} and \ce{N2}, and to some extent for the equilibrium properties of \ce{F2}.
These are exceptions though, as the correction virtually always improve the accuracy of the computed observables.

@@

Including the EN2 correction thus accelerates the rate of convergence as a function of $\Ndet$, when compared to the unperturbed calculations.
In fact, the improvement is such that a given CI+EN2 model often provides more accurate results than the higher-order bare CI model sharing the same computational scaling.
For instance, hCI1+EN2 and hCI2 display the same $\order*{N^6}$ computational scaling, and the former outperforms the latter in many cases.
In Figs.~\ref{fig:plot_all} and ~\ref{fig:plot_all_closed}, hCI1+EN2 correlates with the second light green mark, and hCI2 with the fourth dark green mark.
The former model provides smaller NPEs and distance errors for \ce{OH}, \ce{HF}, \ce{F2} (all single bond breaking),
the opposite being true for multiple bond breaking.
The advantage of the EN2 correction is more striking for the equilibrium properties,
where hCI1+EN2 is more accurate than hCI2 for \ce{OH}, vinyl, ethylene, and \ce{N2},
whereas both models perform similarly for \ce{HF}, \ce{F2}, \ce{H4}, \ce{H7}, and \ce{H8}.

A similar trend on the effect of the EN2 correction also holds for excitation-based CI.
For that, we compare CISD+EN2 and CISDT, both sharing a $\order*{N^8}$ scaling.
CISD+EN2 correlates with the third dark red mark and CISDT with the fourth light red mark in Figs.~\ref{fig:plot_all} and ~\ref{fig:plot_all_closed}.
The errors for the four observables are systematically smaller with the former model,
except for the NPE and distance error of \ce{H8} and the equilibrium properties of \ce{H7}, where they are comparable.
%\ce{OH}, \ce{HF}, \ce{F2}, ethylene, \ce{N2}, \ce{H4} (small)

Importantly, the hCI+EN2 route outperforms its excitation-based counterpart, thus preserving the advantage displayed without the perturbative correction.
Furthermore, the convergence of the observables is overall more monotonic with the EN2 correction.
In this sense, it regularizes oscillations seen in the unperturbed case.
This is clearly seen in the equilibrium properties of vinyl, \ce{H7}, and \ce{F2},
and in the NPE of \ce{OH}, \ce{HF}, \ce{F2}, and ethylene.

The EN2 correction usually ameliorates the performance of the sCI route, though much less observed for hCI and excitation-based CI.
This reflects the poorer reference provided by sCI, from which perturbation theory has a hard time to recover from.
In a handful of cases, however, the EN2 correction does not improve the bare sCI results,
for both NPEs (\ce{OH}, \ce{N2}, \ce{F2}, and ethylene)
and equilibrium properties (vinyl, \ce{HF}, \ce{F2}, and ethylene).

@@

For each CI model considered here, we evaluated the mean signed error (MSE), mean absolute error (MAE), and root-mean square error (RMSE) 
with respect to reference theoretical values for the excitation energies.
The statistical errors are presented in Tab.~\ref{tab:stat1} and Tab.~\ref{tab:stat2},
for singly excited states of closed-shell and open-shell systems, respectively.
For the lower order models, the calculations were performed for 48 (closed-shell) and 19 (open-shell) excited states.
In turn, a subset of 16 (closed-shell) and 6 (open-shell) excited states was considered for the higher order models, given their more intensive computational cost.
Even though these subsets are too small for meaningful absolute statistics, they can still show the main trends on the higher order CI models.
%A detailed comparison of excitation-based CI along the ground-state and the state-specific routes can be found elsewhere. \cite{Kossoski_2023}
%The focus of the present discussion lies on the comparison between hCI, sCI, and excitation-based CI, for both routes,

@@

We start the discussion with the excitations from closed-shell systems, with corresponding statistical errors shown in Tab.~\ref{tab:stat1}.
%
The first level of hCI, hCI1, produces poor excitation energies, with a MAE of \SI{1.07}{\eV}.
This model shares the same computational scaling with CIS, but includes the paired double excitations, 
which clearly do not improve the descent CIS results (MAE of \SI{0.61}{\eV}).
Moving one rank up, to hCI1.5, the MAE increases to \SI{1.91}{\eV}, and reaches a maxium of \SI{3.53}{\eV} at the hCI2 level.
Here we notice that hCI2 delivers somewhat smaller errors than CISD (MAE of \SI{4.09}{\eV}), even though both are way too large.
From that point on, the next hCI models generate progressively better results.
Despite the improvement with respect to hCI2, hCI2.5 is stil as inaccurate (MAE of \SI{1.95}{\eV}) as hCI1.5.
At the hCI3 level, significantly smaller errors are finally achieved (MAE of \SI{0.19}{\eV}).
This is quite close to that obtained with the excitation-based model of same order, CISDT (\SI{0.17}{\eV}).
Even smaller errors are produced at the hCI3.5 level (MAE of \SI{0.11}{\eV}), but at a considerable computational cost.

@@

%we further investigated whether higher-order versions could be competitive against the traditional excitation-based counterparts.
%For that, we performed calculations for a subset of 16 singly-excited states of small closed-shell molecular systems (indicated in the {\SupInf}).
%(\ce{H2O}, \ce{BF}, \ce{HCF}, \ce{HCl}, and \ce{N2}),
%with corresponding statistical errors shown in Tab.~\ref{tab:stat1}.
%Even though this is a small set, it should be enough to obtain the main trends on the accuracy of higher-order models.

@@

This residual bias disappears at the next order, $\Delta$hCI2, whereas for the closed-shell systems even $\Delta$hCI3.5 still understimated excitation energies. 
The same holds for excitation-based CI.
For open-shells, $\Delta$hCI2 is arguably slightly more accurate (MAE of \SI{0.11}{\eV}) than $\Delta$CISD (MAE of \SI{0.12}{\eV}), also based on the larger differences on the RMSEs.
In further contrast to the closed-shell case,
$\Delta$hCI2.5 is more accurate than $\Delta$hCI2, producing a MAE of \SI{0.06}{\eV}.
(Accounting only for the 6 states considered for $\Delta$hCI2.5, $\Delta$hCI2 would have a MAE of \SI{0.09}{\eV}.)
The statistics at the next orders ($h=3$ and $h=3.5$) is limited, 
but suggests a possible advantage of the hCI models against the excitation-based ones,
and that the ground-state routes perform slightly better than the state-specific ones.

\begin{table}[ht!]
\caption{Mean Signed Error (MSE), Mean Absolute Error (MAE), and Root-Mean Square Error (RMSE) in Units of eV, with Respect to Reference Theoretical Values, for the Set of 
Singly-Excited States from Open-Shell Doublets Listed in the {\SupInf}.
}
\label{tab:stat2}
\begin{ruledtabular}
\begin{tabular}{ldddd}
method            & \mc{1}{c}{count} & \mc{1}{c}{MSE} & \mc{1}{c}{MAE} & \mc{1}{c}{RMSE} \\
\hline
CIS               & 19 & +0.33 & 0.36 & 0.50 \\
CISD              & 6  & +1.98 & 1.98 & 2.16 \\
CISDT             & 6  & +0.02 & 0.03 & 0.05 \\
CISDTQ            & 6  & +0.03 & 0.03 & 0.05 \\
%
\hline
hCI1              & 19 & +0.70 & 0.77 & 0.96 \\
hCI1.5            & 19 & +1.11 & 1.11 & 1.30 \\
hCI2              & 6  & +0.69 & 0.69 & 0.85 \\
hCI2.5            & 6  & +0.01 & 0.03 & 0.04 \\
hCI3              & 6  & -0.00 & 0.02 & 0.03 \\
hCI3.5            & 6  & -0.00 & 0.01 & 0.02 \\
%hCI4              & 6  &       &      &      \\
%
\hline
$\Delta$CSF       & 19 & -0.04 & 0.41 & 0.58 \\
$\Delta$CISD      & 19 & +0.01 & 0.12 & 0.21 \\
$\Delta$CISDT     & 6  & -0.05 & 0.05 & 0.09 \\
$\Delta$CISDTQ    & 6  & -0.02 & 0.02 & 0.02 \\
%
\hline
$\Delta$hCI1      & 19 & -0.19 & 0.34 & 0.48 \\
$\Delta$hCI1.5    & 19 & -0.10 & 0.46 & 0.70 \\
$\Delta$hCI2      & 19 & -0.01 & 0.11 & 0.16 \\
$\Delta$hCI2.5    & 6  & -0.06 & 0.06 & 0.12 \\
$\Delta$hCI3      & 6  & -0.03 & 0.03 & 0.06 \\
$\Delta$hCI3.5    & 6  & -0.02 & 0.02 & 0.02 \\
%$\Delta$hCI4      & 6  &       &      &      \\
%
\hline
sCI1              & 19 & +0.24 & 0.39 & 0.61 \\
$\Delta$sCI1      & 19 & -0.11 & 0.27 & 0.38 \\
\end{tabular}
\end{ruledtabular}
\end{table}
